Slovene NLTK Tagger

### 2012-01-15 paper/paper.tex
Corrected some mistakes.

The Brill tagger is another part-of-speech tagger and was first described by Eric Brill in 1993. Before trying to explain it, we must address issues with n-gram taggers (e.g. \textit{Trigram tagger}).

First issue with n-gram taggers is size of their n-gram table. If we want to use tagging process on mobile computing device, compromise must be taken between performance and model size.

A second issue with n-gram tagger is, that the only information it considers from prior context is tags. This is not practical because words themselves are useful information also. Therefore n-gram taggers are impractical models to be used to identify words in context.

To bypass those issues we can use \textit{Brill tagger}. It can be summarised as an transformation-based learning and performs very well using models that are just a fraction of the size of n-gram taggers.

Despite obvious differences between Brill tagger and n-gram tagger there is one similarity. Both of them use \textit{supervised learning} method, since we require annotated training data to make sure whether the taggers guess is a mistake or not.

@@

\subsection[Naive Bayes classifier]{Naive Bayes classifier\footnote{Definition and images summarized from \cite{NLTKBOOK}}}
Both Brill and Trigram are sequential backoff taggers. Meanwhile, Naive Bayes is \textit{classifier based tagger}.
Classification is a process where, for a given output, we have to choose the correct class. Basically, it is an  procedure for assigning input data into one of a given number of categories.

Examples of classification tasks are:
\begin{itemize}
\item whether an email is spam or not.
\item diagnosis to a given patient as described by observed characteristics of the patient.
\end{itemize}
The basic classification task has a number of interesting variants. For example, in multi-class classification, each instance may be assigned multiple labels; in open-class classification, the set of labels is not defined in advance; and in sequence classification, a list of inputs are jointly classified.

@@

Let's assume that in the training corpus most documents have automotive topic. Because of this assumption, the classifier starts near "automotive" label. Later it must take into consideration the effect of each feature. Example considers, that input document contains word "dark", which indicates for "murder mystery", but also contains the word "football", which (strongly) indicate for "sports". After every feature has made its contribution, the classifier checks which label it is closest to, and assigns that label to the input.

@@

\section{Tools used}
\label{toolsUsed}
\subsection{Nltk-trainer} %NIKICC
The code for building the tagger was taken from \textit{Nltk-trainer}  project \cite{nltk-trainer}, whose author is Jacob Perkins.
Its an open-source project hosting on Github, with moto to \textit{Train NLTK objects with zero code}.
We almost exclusively used the script \texttt{train\_tagger.py}, which has the ability to generate NLTK taggers. This project made it a lot easier for us to train the taggers on Slovenian corpora.

\subsection{JOS corpus} %BANIČ
The basic component for building Slovene tagger is \textit{JOS corpus} from JOS project \cite{JOS}. It stands for: \textit{Jezikoslovno Označevanje Slovenskega jezika} (Linguistic annotation of Slovenian language).
It contains collections of various text in Slovenian language with annotation.
For our project we used jos1M corpus, that contain 1 million words with it's appropriate lemmas and morphosyntactic descriptions.
The JOS is compatible with Slovene\textit{ MULTEXT-East morphosyntactic specifications Version 4}\cite{MULTEXT-East}.
It basic purpose is to define word classes (i.e. part-of-speech).
For each class there are various attributes and their appropriate values, which they can be mapped into morphosyntactic descriptions (i.e. MSD).
The structure of JOS corpora is in  \textit{Extensible Markup Language (XML)}.

@@

The trainer will train the tagger properly, if .pos file consists off:
\begin{itemize}
\item The words and their MSD are separated with slash.
\item Every punctuation mark is separated with slash and then follows the same punctuation mark.
\item Words and punctuation marks are between space.
\end{itemize}

If we follow these "rules", implementation is simple. We just need to go through all child tags in sentence tag and properly write them in a file. Because Slovenian language contain letters with caron, proper encoding of tags must be provided. This is simply done with function \textit{encode()} (e.g. encode("utf-8") - this is proper encoding for Slovenian language).
\textit{Jos1M} corpus is build up with 10 XML files (e.g. \texttt{jos1M-01.xml},  \texttt{jos1M-02.xml}, ..., \texttt{jos1M-10.xml}), so we must parse all of them. 
File \texttt{transformJOS.py} contains the corpus transformation.

@@

\subsection{Usage with NLTK} %BANIČ

We can use the trained tagger and NLTK to see how they can be used in Slovenian language. The requirements are installed NLTK and Python 2.7. We set an example in file \texttt{example.py}.
First we need to import following functions:
\begin{itemize}
\item[•] \textit{pickle} - it is a module for serialization and de-serialization a Python object structure ("pickling" - Python object hierarchy is converted into a byte stream). Pickle is used for loading classifiers such as Trigram tagger, Brill tagger, etc.
\begin{lstlisting}
tagger = pickle.load( open("slovene_taggers/BrillTagger.pickle") )
\end{lstlisting}

\item[•] \textit{PunktWordTokenizer} \cite{tokenizer} - it is just one of many NLTK's tokenizers, that can be used to divide strings into lists of substring. Tokenizers are subclasses of \textit{nltk.tokenizer.TokenizerI} interface, which defines the \textit{tokenizer()} method. PunktWordTokenizer uses regular expressions to divide a word into tokens. These tokens are then used by the tagger. Following example will demonstrate, how does PunktWordTokenizer tokenize a sentence. If we have the following sentence:

@@

[\textit{'Potem', 'se', 'je', 'obrnil', 'proti', 'nam', 'in', 'skozi', 'solze', 'izoblikoval', 'najlepši', 'nasmeh', ',', 'kar', 'sem', 'jih', 'videl', 'v', 'življenju.'}]

As you can see, there is a problem with the last token. The solution we used is described below.

\item[•] \textit{nltk.data} - it is a module that contains various functions to load NLTK files, such as corpora. We use it for \textit{sentence segmentation}. In some cases user can input not a sentence, but a whole paragraph. Before using PunktWordTokenizer to tokenize sentence into tokens, we need to segment it into sentences. This can be easily used by including Slovene Punkt sentence segmenter. Following example shows, how does sentence segmenter work. If we have the following paragraph:

@@

the sentence segmenter will output the following list of sentences:

@@

The first step is to use sentence segmenter, to form a list of sentences. This list is then being tokenized with PunktWordTokenizer. One drawback with this tokenizer is that it doesn't tokenize the last word of a sentence and the following period (as shown in the example). Before using the PunktWordTokenizer, we must divide them with space. After we tokenize the sentences, tagging can be used. The result is a list of tuples with words or characters and their appropriate morphosyntactic description.

We also implemented a function that returns two description dictionaries for MSD's with attribute=value expansions in Slovene and English. This way user can view which part-of-speech words belong to in Slovene or English. Following example demonstrates the expected output:
\\\\
\textit{( Lep  |  PPNMEIN ) - pridevnik \\
	( je  |  GP-STE-N ) - glagol\\
	( dan  |  SOMEI ) - samostalnik\\
	( ,  |  , ) - ni razlage\\
	( vse  |  ZC-SEI ) - zaimek\\
	( diši  |  GGNSTE ) - glagol\\
	( že  |  L ) - členek\\
	( po  |  DM ) - predlog\\
	( pomladi  |  SOZEM ) - samostalnik\\
	( !  |  ! ) - ni razlage}\\

We must exmplain, that here only the part of whole MSD is printed. For example the whole MSD for \textit{Lep} shown in table \ref{tab:MSDLep}.

@@

The first output was created with function \textit{prettyPrintWithDescription}, full output can be created with \textit{prettyPrintWithFullDescription}. Both function are defined in \texttt{example.py}
\par
But there is an issue. Because Slovenian language contain letters with caron, the output will have unicode symbols for them. In the previous examples, the output is already corrected. Solution is described in the following section.

\subsection{Encoding problems} %BANIČ
Printing the result of our example is not optimal. Because we want users to know what these characters are, we implemented a solution. To resolve this, we must decode every word to \textit{UTF - 8} (because words we tag are in UTF - 8). If a user wants to print the result in "pretty" form, he should call function \textit{prettyPrint()} or \textit{prettyPrintByLine()} (both are implemented in \texttt{example.py}).

@@

\section{Conclusion} %BANIČ
To conclude our paper, tagging in Slovene language is practical with proper use of Natural Language Toolkit and \textit{nltk-trainer}. Time spent in seconds for tagging different numbers of words, show that Brill and Trigram ran at the same speed, while Naive Bayes ran drastically slower. There are no major differences in tagging accuracy. However Brill and the Naive Bayes are about 2\% better than Trigram.
\par
Besides these three taggers we built, others can be built without any bigger modification. All that is needed is to set some arguments when calling the script \texttt{train\_tagger.py}. The main reason we built only these three is computing power. These three are the only one we could built and evaluate in some reasonable time.
\par
Our whole project is publicly available on github \cite{CODE}. We also contacted the NLTK developers group and will arrange our taggers to become the part of NLTK.

>>>>

First problem with n-gram taggers is the size of their n-gram table. If we want to use tagging process on mobile computing device, compromise must be taken between performance and model size.

Second problem with n-gram tagger is, that the only information it considers from the prior context are tags. This is not practical because words themselves are useful information. Therefore n-gram taggers are impractical models for identification of words in context.

To bypass those problems we can use \textit{Brill tagger}. The Brill tagger was first described by Eric Brill in 1993. It can be summarised as an transformation-based learning and performs very well using models that are just a fraction of the size of n-gram taggers.

Despite obvious differences between Brill tagger and n-gram tagger there is one similarity. Both of them use \textit{supervised learning} method, since we require annotated training data to check whether the taggers guess is a mistake or not.

@@

\subsection{Naive Bayes classifier}
Both Brill and Trigram are sequential backoff taggers. Meanwhile, Naive Bayes is a \textit{classifier based tagger}.
Classification is a process where, for a given output, we have to choose the correct class. Basically, it is  procedure for assigning input data into one of a given categories.

Examples of classification tasks are:
\begin{itemize}
\item is an email a spam or not.
\item diagnosis of a given disease as described by observed characteristics of the patient.
\end{itemize}
The basic classification task has a number of interesting variants. For example, in multi-class classification, each instance may be assigned multiple labels; in open-class classification, the set of labels is not defined in advance; and in sequence classification, a list of inputs are jointly classified.

@@

Let's assume that in the training corpus most documents have automotive topic. Because of this assumption, the classifier starts near "automotive" label. Later it must take into consideration the effect of each feature. Example considers, that input document contains word "dark", which indicates to "murder mystery", but also contains the word "football", which (strongly) indicates to "sports". After every feature has made its contribution, the classifier checks which label it is closest to, and assigns that label to the input.

@@

Definitions and images in this section are summarized from \cite{NLTKBOOK}.
\section{Tools used}
\label{toolsUsed}
\subsection{Nltk-trainer} %NIKICC
The code for building the tagger was taken from \textit{Nltk-trainer}  project \cite{nltk-trainer}, whose author is Jacob Perkins.
Its an open-source project hosting on Github, with moto to \textit{Train NLTK objects with zero code}.
We almost exclusively used the script \texttt{train\_tagger.py}, which has the ability to generate NLTK taggers. This project made it a lot easier for us to train the taggers on Slovenian corpora.

\subsection{JOS corpus} %BANIČ
The basic component for building Slovene tagger is \textit{JOS corpus} from JOS project \cite{JOS}. It stands for: \textit{Jezikoslovno Označevanje Slovenskega jezika} (Linguistic annotation of Slovenian language).
It contains collections of various text in Slovenian language with annotations.
For our project we used jos1M corpus,  containing 1 million words with appropriate lemmas and morphosyntactic descriptions.
The JOS is compatible with Slovene\textit{ MULTEXT-East morphosyntactic specifications Version 4}\cite{MULTEXT-East}.
Its basic purpose is to define word classes (i.e. part-of-speech).
For each class there are various attributes and their appropriate values, which can be mapped into morphosyntactic descriptions (i.e. MSD).
The structure of JOS corpora is described in a form of \textit{Extensible Markup Language (XML)}.

@@

The trainer will train the tagger properly, if .pos file consists off:
\begin{itemize}
\item The words and their MSD are separated with slash.
\item Every punctuation mark is separated with slash and then follows the same punctuation mark.
\item Words and punctuation marks are separated with spaces.
\end{itemize}

If we follow these "rules", implementation is simple. We go through all child tags in a sentence tag and write them to a file. Because Slovenian language contain letters with caron, proper encoding of tags must be provided. This is done with function \textit{encode()} (e.g. encode("utf-8") - for Slovenian language).
\textit{Jos1M} corpus is build from 10 XML files (e.g. \texttt{jos1M-01.xml},  \texttt{jos1M-02.xml}, ..., \texttt{jos1M-10.xml}), so we parse all of them. 
File \texttt{transformJOS.py} contains the corpus transformation.

@@

\subsection{Use of taggers with NLTK} %BANIČ

We can apply the trained tagger and NLTK to see how they can be used in Slovenian language. The pre-requirements are NLTK and Python 2.7. We present an example in file \texttt{example.py}.
First we need to import the functions:
\begin{itemize}
\item[•] \textit{pickle} - it is a module for serialization and de-serialization a Python object structure ("pickling" - Python object hierarchy is converted into a byte stream). Pickle is used for loading classifiers such as Trigram tagger, Brill tagger, etc.
\begin{lstlisting}
tagger = pickle.load( open("slovene_taggers/BrillTagger.pickle") )
\end{lstlisting}

\item[•] \textit{PunktWordTokenizer} \cite{tokenizer} - it is just one of many NLTK's tokenizers, that can be used to divide strings into lists of substring. Tokenizers are subclasses of \textit{nltk.tokenizer.TokenizerI} interface, which defines the \textit{tokenizer()} method. PunktWordTokenizer uses regular expressions to divide a word into tokens. These tokens are then used by the tagger.  If we have the following sentence:

@@

[\textit{'Potem', 'se', 'je', 'obrnil', 'proti', 'nam', 'in', 'skozi', 'solze', 'izoblikoval', 'najlepši', 'nasmeh', ',', 'kar', 'sem', 'jih', 'videl', 'v', 'življenju.'}]
\par

As you can see, there is a problem with the last token, because the tokenizer did not separate \textit{življenju} and period. The tagger will not recognize [\textit{življenju.}], because it is not in a valid form (i.e. words and punctuation marks must be separated tokens). Before tagging we must separate this type of tokens into two new tokens.

\item[•] \textit{nltk.data} - it is a module that contains various functions to load NLTK files, such as corpora. We use it for \textit{sentence segmentation}. In some cases the user can input a whole paragraph. Before using PunktWordTokenizer to tokenize sentence into tokens, we need to segment it into sentences. This can be easily done by including Slovene Punkt sentence segmenter. The example shows, how sentence segmenter works:

@@

The sentence segmenter will output the following list of sentences:

@@

The first step is to use sentence segmenter, to form a list of sentences. This list is then tokenized with PunktWordTokenizer. A drawback is that the last word of a sentence and the following period (as shown in the example) are not tokenized. Before using the PunktWordTokenizer, we must divide them with space. After we tokenize the sentences, tagging can be used. The result is a tuple list with words or characters and their morphosyntactic description.

We also implemented a function that returns two description dictionaries for MSD's with attribute=value expansions in Slovene and English. Thereby a user can check which part-of-speech words belong to Slovene or English. The expected output is:
\\\\
\textit{( Lep  |  PPNMEIN ) - pridevnik \\
	( je  |  GP-STE-N ) - glagol\\
	( dan  |  SOMEI ) - samostalnik\\
	( ,  |  , ) - ni razlage\\
	( vse  |  ZC-SEI ) - zaimek\\
	( diši  |  GGNSTE ) - glagol\\
	( že  |  L ) - členek\\
	( po  |  DM ) - predlog\\
	( pomladi  |  SOZEM ) - samostalnik\\
	( !  |  ! ) - ni razlage}\\

Here only a part of the whole MSD is presented. The whole MSD for \textit{Lep} is shown in table \ref{tab:MSDLep}.

@@

The first output was created with function \textit{prettyPrintWithDescription}, full output can be created with \textit{prettyPrintWithFullDescription}. Both function are defined in \texttt{example.py}.
\par
Because Slovenian language contain letters with caron, the output will have unicode symbols for them. In the previous examples, the output is already corrected.

Printing the result of our example is not optimal. We decode every word to \textit{UTF - 8} (because words we tag are in UTF - 8). If a user wants to print the result in "pretty" form, he should call function \textit{prettyPrint()} or \textit{prettyPrintByLine()} (both are implemented in \texttt{example.py}).

@@

\section{Conclusion} %BANIČ
To conclude our paper, tagging in Slovene language is practical with proper use of Natural Language Toolkit and \textit{nltk-trainer}. Time for tagging show that Brill and Trigram ran at the same speed, while Naive Bayes is much slower. There are no major differences in tagging accuracy, but Brill and the Naive Bayes are about 2\% better than Trigram.
\par
Besides the taggers we built, others can be built without any large modifications by calling the script \texttt{train\_tagger.py}.
\par
Our whole project is publicly available on github \cite{CODE}. We contacted the NLTK developers group and will arrange our taggers to become a part of NLTK.
